### 2022-09-08 README.Rmd

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/njtierney/conmat/branch/master/graph/badge.svg)](https://codecov.io/gh/njtierney/conmat?branch=master)
[![R-CMD-check](https://github.com/njtierney/conmat/workflows/R-CMD-check/badge.svg)](https://github.com/njtierney/conmat/actions)
<!-- badges: end -->

@@

Contact matrices describe the degree of contact between individuals of given age groups. These matrices are commonly used to model how diseases such as COVID-19 spread in a population through social contact.

@@

```{r get-polymod}
library(conmat)
polymod_contact_data <- get_polymod_setting_data()
polymod_survey_data <- get_polymod_population()
```

The contact data is a list of the contacts for each of the settings, "home", "work", "school", and "other". It contains a dataframe for each of these.

```{r show-polymod-str}
str(polymod_contact_data)
```

Say looking at the "home" setting, each row contains survey information of the number of contacts. Specifically, the number of contacts from one age group to another age group, and then the nuber of participants in that age group.

```{r show-polymod-home}
polymod_contact_data$home
```

The survey data contains the lower age limit and the population in that age group.

@@

We can create a model for each of the settings with the function `fit_setting_contacts`

```{r fit-polymod}
set.seed(2021-09-24)
setting_models <- fit_setting_contacts(
  contact_data_list = polymod_contact_data,
  population = polymod_survey_data
  )
```

This contains a list of models, one for each setting. We can look at one, and get summary information out:

```{r}
names(setting_models)
setting_models$home
```

So this gives us our baseline model of a contact model. We have fit this model to the existing contact survey data.

We can now predict to another age population, to create our "synthetic" contact matrix. Say we want to create a contact matrix for the "Fairfield" LGA, we can extract it out like so:

```{r get-fairfield}
library(conmat)
fairfield_age_pop <- abs_age_lga("Fairfield (C)")
fairfield_age_pop
```

Then we take the model we had earlier, and extrapolate to the fairfield data with `predict_setting_contacts`, also providing some age breaks we want to predict to

```{r fairfield-synth-5}
set.seed(2021-09-24)
synthetic_settings_5y_fairfield <- predict_setting_contacts(
  population = fairfield_age_pop,
  contact_model = setting_models,
  age_breaks = c(seq(0, 85, by = 5), Inf)
  )
```

This then returns a list of synthetic matrices, "home", "work", "school", "other", and "all", which is the sum of all matrices.

```{r}
str(synthetic_settings_5y_fairfield)
synthetic_settings_5y_fairfield$home
synthetic_settings_5y_fairfield$all
```

@@

Looking at the matrices is nice, but let's plot them. We can use `plot_setting_matrix` to plot all at once

```{r fairfield-synth-5-plot}
# this code is erroring for the moment - something to do with rendering a large plot I think.
plot_setting_matrices(
  synthetic_settings_5y_fairfield,
  title = "Setting-specific synthetic contact matrices (fairfield 2020 projected)"
)
```

You can also plot individual matrices with `plot_matrix`

```{r plot-matrix-differents}
plot_matrix(synthetic_settings_5y_fairfield$home)
plot_matrix(synthetic_settings_5y_fairfield$all)
```

## Speeding up computation with `future`

`conmat` supports parallelisation, which is useful in a couple of contexts with the model fitting, here is an example:

```{r load-future}
library(future)
plan(multisession, workers = 4)
```

We set the future plan, saying "multisession", with 4 workers. Then we run the same code as above (note that you must specify the plan, otherwise it does not know how to parallelise. See the [future package documentation](https://future.futureverse.org/reference/plan.html) for more details):

```{r show-off-furrr}
polymod_setting_data <- get_polymod_setting_data()
polymod_population <- get_polymod_population()

contact_model <- fit_setting_contacts(
  contact_data_list = polymod_setting_data,
  population = polymod_population
)

contact_model_pred <- predict_setting_contacts(
  population = polymod_population,
  contact_model = contact_model,
  age_breaks = c(seq(0, 75, by = 5), Inf)
)
```

Notably this is about 3 times faster than without using that plan.

@@

This package provides data and helper functions for the data, for use in calculating contact matrices. The data sources are from the Australian Bureau of Statistics (ABS), as we were using these a lot when we created the package. In the future we might wrap these data sources and helpers into another package, but for the time being they are here.

Below are a couple of examples of data provided, see the "data sources" vignette and helpful at the website for full details.

You can extract the age population structure for the LGA Fairfield, or Brisbane, like so:

```{r abs-age-lga}
abs_age_lga("Fairfield (C)")
abs_age_lga("Brisbane (C)")
```

@@

```{r abs-age-state}
abs_age_state(state_name = "QLD")
abs_age_state(state_name = "NSW")
```

>>>>

<!-- badges: start -->
[![R-CMD-check](https://github.com/njtierney/conmat/workflows/R-CMD-check/badge.svg)](https://github.com/njtierney/conmat/actions)
[![Codecov test coverage](https://codecov.io/gh/njtierney/conmat/branch/master/graph/badge.svg)](https://codecov.io/gh/njtierney/conmat?branch=master)
<!-- badges: end -->

@@

Contact matrices describe the degree of contact between individuals of given age groups.

For example, this matrix describes the number of contacts between individuals

```{r}
#| label: show-contact
#| echo: FALSE

name_vec <- c("0-4", "5-9", "10-14")
cmat <- matrix(
  data = rep(NA, 9),
  nrow = 3,
  ncol = 3,
  byrow = TRUE,
  dimnames = list(
    name_vec,
    name_vec
    )
)

diag(cmat) <- c(10,11,13)
cmat[upper.tri(cmat)] <- 3:5
cmat[lower.tri(cmat)] <- 3:5

cmat
```

The rows and columns represent the age groups of the people. On the main diagonal we see that we have a higher number of contacts - showing that people of similar ages tend to interact more with one another.

We can use the information in these matrices to model how diseases such as COVID-19 spread in a population through social contact.

@@

```{r get-polymod}
library(conmat)
polymod_contact_data <- get_polymod_contact_data(setting = "work")
polymod_survey_data <- get_polymod_population()
```

The contact data is a data frame containing the age from and to, and the number of contacts for each of the specified settings, "home", "work", "school", "other", or "all" as well as the number of participants. By default, `polymod_contact_data` contains data from "all", but we're going to use the "work" set of data, as it produces an interesting looking dataset. Each row contains survey information of the number of contacts. Specifically, the number of contacts from one age group to another age group, and then the number of participants in that age group.

The survey data, `polymod_survey_data` contains the lower age limit and the population in that age group.

@@

## Predicting the contact rate

We can create a model of the contact *rate* with the function `fit_single_contact_model`

```{r fit-polymod}
set.seed(2022-09-06)
contact_model <- fit_single_contact_model(
  contact_data = polymod_contact_data,
  population = polymod_survey_data
  )
```

This fits a generalised additive model (GAM), predicting the contact rate, based on a series of prediction terms that describe various features of the contact rates.

```{r show-conctact-model}
contact_model
```

We can use this contact model to then predict the contact rate in a new population.

As a demonstration, let's take an age population from a given LGA in Australia (this was the initial motivation for the package, so there are some helper functions for Australian specific data).

```{r fairfield}
fairfield_age_pop <- abs_age_lga("Fairfield (C)")
fairfield_age_pop
```

We can then pass the contact model through to `predict_contacts`, along with the fairfield age population data, and some age breaks that we want to predict to.

```{r predict-contacts}
set.seed(2022-09-06)
synthetic_contact_fairfield <- predict_contacts(
  model = contact_model,
  population = fairfield_age_pop,
  age_breaks = c(seq(0, 85, by = 5), Inf)
)

synthetic_contact_fairfield
```

@@

Let's visualise the matrix to get a sense of the predictions with `plot_matrix`. First we need to transform the predictions to a matrix:

```{r plot-matrix-differents}
synthetic_contact_fairfield %>% 
  predictions_to_matrix() %>% 
  plot_matrix()
```

## Applying the model across all settings.

You can also fit a model for all of the settings all at once with a series of functions, `fit_setting_contacts`, and `predict_setting_contacts`.  This means we can do the above, but for each setting, "home", "work", "school", "other", and "all". We would recommend this when using conmat, as it is a pretty common use case. However for demonstration purposes we wanted to show how it works for a single matrix here first. We also provide details on how to fit the model to each of these settings in parallel. For more details on that workflow, see the "getting started" vignette.

@@

This package provides data and helper functions for the data, for use in calculating contact matrices. The data sources are from the Australian Bureau of Statistics (ABS), as we were using these a lot when we created the package. In the future we might wrap these data sources and helpers into another package, but for the time being they are here. Below are a couple of examples of data provided, see the "data sources" vignette and helpful at the website for full details.

You can extract the age population structure for the LGA, Brisbane, like so:

```{r abs-age-lga}
abs_age_lga("Brisbane (C)")
```

@@

```{r abs-age-state}
abs_age_state(state_name = "QLD")
```
